\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-01-10 pea_paper_docs/tex/pea.tex
modified hypergeometric scripts
plotting is difficult since I must remove WB identifiers, yet these are
different lengths depending on the ontology.

I added an ‘analysis’ variable to the plot_enrichment and
fetch_dictionary functions to let the user specify what analysis is
performed.

@@

\section*{Methods}
\subsection*{Building the dictionaries}

\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{} to extract information on all genome-wide
association studies deposited there. We only selected traits that had $>300$
associated genes. We identified 24 traits that met our criteria. Next, we used
DIOPT~\cite{} to identify candidate homologs for the genes associated with these
traits. Briefly, DIOPT combines a large number of methods for identifying homologs
and returns homolog candidates associated with a compound score. Depending on the score,
homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
in the homology. Many-to-one and one-to-many homology relationships are allowed
in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
For our study, we only accepted homolog candidates with `high' or `moderate' scores
and we did not insist on a one-to-one relationship between genes.

@@

@@

\subsection*{Deconstructing the Egl phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the egg-laying defective (Egl) phenotype.
In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
in the egg-laying control~\cite{}. The complexity of the interactions that
happen to allow egg-laying make understanding the Egl phenotype in terms of
tissues a challenging task.

@@

\section*{Conclusions}
The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
an important step towards a unified set of analyses that can help researchers
to understand genomic datasets. By offering tissue, phenotype and gene ontology
enrichment on a single site, researchers will benefit both in the speed at which
the analysis is completed, but also from the fact that the analyses are all
carried out with the same methodology and the same underlying algorithms.

>>>>

% more space between rows
\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}

@@

\section*{Methods}
\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{} to extract information on all genome-wide
association studies deposited there. We only selected traits that had $>300$
associated genes. We identified 24 traits that met our criteria. Next, we used
DIOPT~\cite{} to identify candidate homologs for the genes associated with these
traits. Briefly, DIOPT combines a large number of methods for identifying homologs
and returns homolog candidates associated with a compound score. Depending on the score,
homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
in the homology. Many-to-one and one-to-many homology relationships are allowed
in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
For our study, we only accepted homolog candidates with `high' or `moderate' scores
and we did not insist on a one-to-one relationship between genes.

@@

% TODO: SO.... ?

@@

\subsection*{Deconstructing phenotype-tissue relationships}
\subsubsection*{Tissue enrichment on the Egl gene set reveals cellular
components of the phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the egg-laying defective (Egl) phenotype.
In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
in the egg-laying control~\cite{}. The complexity of the interactions that
happen to allow egg-laying make understanding the Egl phenotype in terms of
tissues a challenging task.

@@

\subsubsection*{Quantifying the anatomy-phenotype
                mapping via Bayesian probabilities}
Another way to understand the phenotype-anatomy mapping is by considering
how informative a given anatomy term is on a particular phenotype, or vice-versa.
To this end, we calculated two conditional probabilities that helped us answer
this question. The first conditional probability,

\begin{equation}
  P( \text{a gene has Egl annotation} | \text{it is expressed in } X)
\end{equation}

answers the question: For a gene with an expression pattern that includes the
tissue term X (i.e., the gene is expressed at least in X), what is the probability
that this gene has an Egl phenotype (i.e., the phenotype annotations for this
gene include Egl)? For simplicity, we can re-write this equation more succintly
by removing a few words. The calculation of this probability is straightforward
and follows from the definition of conditional probability:

\begin{equation}
  P(\text{Egl}|X) = \frac{N_\text{genes annotated Egl and X}}
                         {N_\text{genes annotated with X}}.
\label{egl_x}
\end{equation}

Equation~\ref{egl_x} measures how likely a gene is to be annotated with an Egl
phenotype given that its expression pattern includes the term $X$. A related
quantity (which is neither the inverse nor the complement) is the conditional
probability that a gene which is annotated with at least the Egl phenotype is
expressed in tissue $X$. That is to say,

\begin{equation}
  P(X|\text{Egl}) = \frac{N_\text{genes annotated Egl and X}}
                         {N_\text{genes annotated with Egl}}.
  \label{x_egl}
\end{equation}

Equation~\ref{x_egl} tells us how probable it is that any given gene that is
annotated with an Egl phenotype includes $X$ as a tissue term. Taken together,
equations~\ref{egl_x} and~\ref{x_egl} help us understand how predictive
anatomic expression is of phenotypes, and how predictive phenotypes are of
anatomic expression.

\begin{table*}
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering{}
  \ra{1.3}
  \caption{Conditional probabilities for various tissues. The
  first column shows the conditional probability that a gene has an Egl
  phenotype given that it has expression in tissue $X$ (given by the row).
  The second column shows the conditional probability that a gene has expression
  in the anatomy term X given that it has an Egl phenotype. The first 9 terms
  are the terms for which $P(\text{Egl}|X)$ is maximized. The last three terms
  are the terms which have the highest $P(X|\text{Egl})$. For clarity, the Pn.p
  cells are not shown even though $P(\text{Egl}|\text{Pn.p})\sim 0.24$.}

  \begin{tabular}{@{}lcc@{}}
  \toprule{}
  Tissue & $P(\text{Egl}| X)$ & $P(X|\text{Egl})$\\
  \bottomrule{}\\
  % Data goes here
  P7.pa & $0.30$ & $0.04$\\
  HSN & $0.27$ & $0.11$\\
  vulC & $0.27$ & $0.06$\\
  vulD & $0.26$ & $0.07$\\
  vulE & $0.25$ & $0.06$\\
  vulF & $0.24$ & $0.06$\\
  vulA & $0.24$ & $0.05$\\
  vulB2 & $0.23$ & $0.05$\\
  vulB1 & $0.22$ & $0.05$\\
  nervous system & $0.02$ & $0.72$ \\
  pharynx  & $0.00$ & $0.46$\\
  sex organ & $0.07$ & $0.41$\\
  tail & $0.05$ & $0.33$\\
  \bottomrule{}
  \end{tabular}
\label{tab:cond_probs}
\end{table*}

We calculated the conditional probability that a gene has an Egl phenotype given
that it's expression pattern includes a tissue term $X$ and we searched for the
tissue terms that maximized this probability. The list of terms that maximized this
probability reflected the results from running TEA on the subset of genes that have
an Egl phenotype. We also calculated the conditional probability that a gene has
expression in a tissue term $X$ given that it is annotated with an Egl phenotype
and we searched for terms that maximized this probability. The terms that maximized
this probability were `nervous system', `pharynx' (a body part with a lot of neurons),
`sex organ' and `tail' (a body part with neurons and hypodermis). In general, the
terms that had a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$.
Additionally, the terms that had a high $P(X|\text{Egl})$ are broad terms that
include a lot of cells, whereas the terms that had a high $P(\text{Egl}|X)$
were considerably more specific.
We conclude that the Egl phenotype arises from a small set of tissues. The Egl
phenotype can be best predicted by genes with expression patterns that include
at least one of a small number of cells (mainly vul cells, HSN). On the other
hand, answering whether the expression pattern of a gene includes a particular
anatomic region or tissue given that the gene has an Egl phenotype is hard to
do for small tissues or single cells. However, guesses about what functional system
or broad anatomic region may be affected by an Egl gene can be answered with
confidence ($\sim70\%$ of the time, the nervous system is the system
affected by an Egl mutant).

\section*{Conclusions}

We have highlighted three possible uses for phenotype and GO enrichment analyses.
Taken together, the WormBase Enrichment Suite can help guide researchers as they
search for a phenologue that is representative of a human disease that has no
immediately obvious counterpart in the worm. Such phenologues may benefit from
the fact that they represent an unbiased approach to disease modeling as long as
the human genes are selected from unbiased screens. GWAS data suggests that certain
aspects of lupus biology may be best understood from a developmental perspective
in \cel{}. On the other hand, our results suggest that rheumatoid arthritis
might not be effectively modeled in the worm, although it is possible that our
inability to find phenologues more concretely is the result of the GWAS screens:
If GWAS screens identified factors that lead to worse prognosis of rheumatoid
arthritis, but not factors that cause onset of the disease, this could explain
the lack of phenotype enrichment in \cel{}.

% Phenologues

% Screen Design

% Curation and annotation biases

The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
an important step towards a unified set of analyses that can help researchers
to understand genomic datasets. By offering tissue, phenotype and gene ontology
enrichment on a single site, researchers will benefit both in the speed at which
the analysis is completed and from the fact that the analyses are all
carried out with the same methodology and the same underlying algorithms.
